Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
Attempt to reformulate the abstract...

\begin{abstract}
XXX the first paragraph is a bit too long, the second too strange, but I think
it is a step in the right direction.

Tracing JIT compilers can greatly speed up programs that spend most of their time
in loops in which they take similar code paths.  Applying an unmodified tracing
JIT to an application that is itself a bytecode interpreter results in very
limited or no speedup. While bytecode interpreters spend most of their time in a
loop, \ie the bytecode dispatch, the code paths taken depend fully on the input
bytecode which the interpreter evaluates and thus changes all the time.

In this paper we show how to guide tracing JIT compilers to greatly improve
the speed of bytecode interpreters towards input bytecode. We evaluate our
technique by applying it to both a small example as well as to a full Python 
interpreter. This research has been conducted in the context of the PyPy 
project.

@@

There have been some attempts to do \emph{dynamic partial evaluation}, which is
partial evaluation that defers partial evaluation completely to runtime
to make partial evaluation more useful for dynamic languages. This concept was
introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
small dynamic language based on lambda-calculus. It is also related to
Psyco \cite{rigo_representation-based_2004}, a specializing JIT compiler for
Python.  There is some work by one of
the authors to implement a dynamic partial evaluator for Prolog
\cite{carl_friedrich_bolz_automatic_2008}. There are also experiments within the
PyPy project to use dynamic partial evaluation for automatically generating JIT
compilers out of interpreters \cite{armin_rigo_jit_2007}\cite{antocuni_2009}.
So far those have not been as
successful as we would like and it seems likely that they will be supplanted
with the work on tracing JITs described here.

>>>>

\begin{abstract}

We attempt to use the technique of Tracing JIT Compilers
(see e.g.\ \cite{andreas_gal_incremental_2006}) in the context
of the PyPy project, \ie on programs that are interpreters for some
dynamic language (including Python).  Tracing JIT compilers can greatly
speed up programs that spend most of their time in loops in which they
take similar code paths.  However, applying an unmodified tracing JIT to
a program that is itself a bytecode interpreter results in very limited
or no speedup.  One of the reasons is that their main loop, \ie the
bytecode dispatch, always executes different bytecodes and so follows
different paths.

In this paper we show how to guide tracing JIT compilers to greatly
improve the speed of bytecode interpreters. We evaluate our technique by
applying it to two PyPy interpreters: one is a small example, and the
other one is the full Python interpreter.

@@

There have been some attempts to do \emph{dynamic partial evaluation}, which is
partial evaluation that defers partial evaluation completely to runtime
to make partial evaluation more useful for dynamic languages. This concept was
introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
small dynamic language based on lambda-calculus. It is also related to
Psyco \cite{rigo_representation-based_2004}, a specializing JIT compiler for
Python.  There is some work by one of
the authors to implement a dynamic partial evaluator for Prolog
\cite{carl_friedrich_bolz_automatic_2008}. There are also experiments within the
PyPy project to use dynamic partial evaluation for automatically generating JIT
compilers out of interpreters \cite{armin_rigo_jit_2007, antocuni_2009}.
So far those have not been as
successful as we would like and it seems likely that they will be supplanted
with the work on tracing JITs described here.
